---
title: "Machine Learning: Evaluate Weight Lifting Techniques"
output: html_document

---

```{r setup, include = FALSE, echo = TRUE, results = "hide"}
ipak <- function(pkg){
      new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
      if (length(new.pkg)) 
            install.packages(new.pkg, dependencies = TRUE)
      sapply(pkg, require, character.only = TRUE)
}

### Package Usage
packages <- c("plyr", "dplyr", "data.table", "dtplyr", "lubridate", "ggplot2", "scales",
                  "reshape2", "knitr", "R.cache", "stringr", "gtools", "quantreg",
                  "graphics", "corrplot", "broom", "rmarkdown", "caret", "randomForest",
                  "gbm", "forecast", "elasticnet", "e1071", "glmnet", "quantmod", "rpart",
                  "rpart.plot", "rattle", "knitr", "tidyverse", "purrr", "mda", "ROCR",
                  "verification", "pROC", "tree", "stringr", "cvTools")
ipak(packages)
```

## Executive Summary  
**Background** 
```r
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  
```
**Analysis Goal**  
```r
In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here:  
```
<http://groupware.les.inf.puc-rio.br/har>  (see the section on the Weight Lifting Exercise Dataset). 

## Data Processing

**Load Data into R**  
```{r load_weightlift_csv, include = TRUE, echo = TRUE, results = "hold"}
train.file = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.file = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(train.file, header = TRUE, 
                  na.strings = c("NA","N/A","NULL","Null","null",""," ","  ","#DIV/0!"))
testing <- read.csv(test.file, header = TRUE, 
                  na.strings = c("NA","N/A","NULL","Null","null",""," ","  ","#DIV/0!"))
```


**Partition Training Data Set**  
To allow for a validation of the models to occur before evaluating predictive accuracy on test set, a validation set, is created.  The validation set will take a portion of the records from the current 'training' set.  The balance of the observations will remain in a training set.  To allow for 2 possible 'cross validation' attempts, the validation set is further split into two data sets, valid1 and valid2. 
```{r testing_partition, include = TRUE, echo = TRUE, results = "markup", cache = TRUE}
set.seed(100)
inTrain = createDataPartition(training$classe, p = 0.6, list = FALSE)
training = training[inTrain,]
validation = training[-inTrain,]

set.seed(100)
inValidation <- createDataPartition(validation$classe, p = 0.5, list = FALSE)
valid1 <- validation[inValidation,]
valid2 <- validation[-inValidation,]
dim(training) 
dim(testing) 
dim(valid1) 
dim(valid2)
```


## Exploratory Data Analysis  
There is a total of 160 variables in the data sets.  To avoid overfitting, we want to identify important or relevant variables, or features, that contribute significantly to greater predictive capability of a model.   A first step is to identify and remove features with a percentage of observations values equal NA or blank is 80% or higher.  The assumption is that the high percentage of NA or blank values raises some concern about the quality of information obtained.  More research is required before further considering these variables for inclusion in the model. This may include better populating values across all observations.
* Step 1: Remove features that have a significant portion of missing / NA values or few unique values (e.g. 1 or 2 different values).  
* Step 2: Run preProcess() with method = "pca" to define principle components that capture at least 80% of variance.  
* Step 3: Plot tree which provides insight to the importance of features considering the interaction between variables.  
* Step 4: Make additional plots to provide further insight to the relationships of each feature, focusing on the principal components/features identified in previous analyses.  
```{r explore_step1, include = TRUE, echo = TRUE, results = "markup"}
pctNA <- sapply(training, function(x) round(sum(is.na(x))/nrow(training),3))
lowNA <- names(pctNA)[pctNA < 0.80]   ## Exclude column names with 80%+ NA
keep <- lowNA[1:length(lowNA)-1]
```
```{r nzv_exclude, include = FALSE, echo = FALSE, eval = FALSE}
## Identify nearZeroVar which likely contribute nominally, if at all, to predictive model
nzv <- nearZeroVar(training, freqCut = 80/20, uniqueCut = 0.1, saveMetrics = TRUE)
nzv.disq <- subset(nzv, zeroVar == TRUE | nzv == TRUE)
nzv.disq <- rownames(nzv.disq)
```


## Model Selection
As our outcome variable 'classe' is qualitative, a linear model does not make sense. We will be using the models listed below to predict the outcome.  In addition to each model indvidually, we will develop a combined, or stacked, model, to see if we can create a materially better model.  If the accuracy of selected model

```{r model_develop, include = TRUE, echo = FALSE, results = "hide", cache = TRUE}
set.seed(100)
fitCtrl <- trainControl(method = "cv", number = 3,repeats = 1)
fitRPart <- train(classe ~ ., data = training[ , c("classe", keep)], method = "rpart", 
                  tuneLength = 10, trControl = fitCtrl, metric = "Accuracy")
predRPart <- predict(fitRPart, newdata = valid1[ ,c("classe", keep)])
confuseRPart <- confusionMatrix(predRPart, valid1$classe)

set.seed(100)
fitCtrl <- trainControl(method = "cv", number = 3,repeats = 1)
fitGBM1 <- train(classe ~ ., data = training[ , c("classe", keep)], method = "gbm", 
                  tuneLength = 10, trControl = fitCtrl, metric = "Accuracy")
predGBM1 <- predict(fitGBM1, newdata = valid1[ ,c("classe", keep)])
confuseGBM1 <- confusionMatrix(predGBM1, valid1$classe)

set.seed(100)
fitCtrl <- trainControl(method = "cv", number = 3,repeats = 1, 
                        classProbs = TRUE, summaryFunction = twoClassSummary)
fitGBM2 <- train(classe ~ ., data = training[ , c("classe", keep)], method = "gbm", 
                  tuneLength = 10, trControl = fitCtrl, metric = "ROC")
predGBM2 <- predict(fitGBM2, newdata = valid1[ ,c("classe", keep)])
confuseGBM2 <- confusionMatrix(predGBM2, valid1$classe)

set.seed(100)
fitCtrl <- trainControl(method = "cv", number = 3,repeats = 1)
fitRF1 <- train(classe ~ ., data = training[ , c("classe", keep)], method = "rf", 
                  tuneLength = 10, trControl = fitCtrl, metric = "Accuracy")
predRF1 <- predict(fitRF1, newdata = valid1[ ,c("classe", keep)])
confuseRF1 <- confusionMatrix(predRF1, valid1$classe)

set.seed(100)
fitCtrl <- trainControl(method = "cv", number = 3,repeats = 1, 
                        classProbs = TRUE, summaryFunction = twoClassSummary)
fitRF2 <- train(classe ~ ., data = training[ , c("classe", keep)], method = "rf", 
                  tuneLength = 10, trControl = fitCtrl, metric = "ROC")
predRF2 <- predict(fitRF2, newdata = valid1[ ,c("classe", keep)])
confuseRF2 <- confusionMatrix(predRF2, valid1$classe)
```


### Exhibit 1: Correlation Factor of Each Variable with MPG
```{r corr_matrix_grid, include = TRUE, echo = TRUE, results = "markup"}
mtcars.cor <- round(cor(mtcars), 3)
mtcars.cor[1,]
```


## Regression Analysis
Here we look build and compare various regression models to identify a model that best fits data.  We use several metrics to evaluate our model fit, including: adjusted r-square, residual squared error (sigma), and p-values.  We also use ANOVA and an analysis of the residuals to evaluate the model. Regression 'fit4' determined to provide best fit.





